A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-12 Manuscript/PNAS-Adams-02.Rmd
vsn 2

The appeal of Pagel's $\lambda$ and Blomberg's $\kappa$ as descriptive statistics is that they provide a basis for interpreting "weak" versus "strong" phylogenetic signal; i.e., small versus large values of $\hat{\lambda}$ or $\kappa$, respectively, in a comparative sense [@DeMeester2019; @Pintanel2019;@Su2019]. Nonetheless, an important question that has yet to be considered is whether these statistics are or can be converted to effect sizes for comparative analyses across datasets? <!-- I think this sentence should be broken into two thoughts: "Nonetheless, an important question that has yet to be considered is whether this comparison is analytically appropriate. To be so, these statistics should be converted to effect sizes to avoid issues of incommensorate units or variation due to different phylogenetic branching patterns." or something like that, identifying why the conversion to effect size is necessary. Otherwise, I think it makes the issue less relatable. I bet few people have considered that question of whether lambda and kappa can be converted to effect sizes, but a much more relatable question is whether people *should* compare the raw values against each other. --> To be statistics representing phylogenetic signal, they should have reliable distributional properties, which could be revealed with simulation experiments. For instance, as a proportional random variable bounded by $0$ and $1$, we might expect that $\hat{\lambda}$ follows a Bernoulli distribution (**add ref**); i.e., branch lengths in a tree are scaled proportionally to the probability that data arise from a BM process. Given a known $\lambda$ value used to generate random data on a tree, we would also expect  that the mean of an empirical sampling distribution of $\hat{\lambda}$ would approximately equal $\lambda$; the dispersion of $\hat{\lambda}$ would be largest at intermediate values of $\lambda$, $\hat{\lambda}$ would be predictable over the range of $\lambda$ with respect to tree size; the distribution of $\hat{\lambda}$ would be symmetric at intermediate values of $\lambda$ and more skewed toward values of 0 or 1; and that the distribution of $\hat{\lambda}$ will be more platykurtic at intermediate values of $\lambda$, becoming more leptokurtic toward 0 and 1 (**add same ref**). Prior work [@Munkemuller_et_al2012] seems to support some of these conjectures, based superficially on statistical moments for a given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of ref. [@Munkemuller_et_al2012]). However, because the "strength of Brownian motion" <!-- unclear what this quote is referring to. Is it in the Munkemuller paper? --> was simulated as a varied weighted-average of data simulated on trees with $\lambda=0$ and $\lambda=1$ and not as prescribed values of $\lambda$ [@Munkemuller_et_al2012], interpretation of these patterns is challenging.

By contrast, for Blomberg's $\kappa$, which is positively uNbounded, we might expect that for any $\lambda$ used to generate data, estimates of $\kappa$ might follow a normal distribution, with values distributed symmetrically about the input value. This attribute seemed less reasonable based on the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], which suggested that distributions were positively skewed and that Blomberg's $\kappa$ might not behave as a statistic that follows a normal distribution. However, because their simulations used a weighted combination of simulated phylogenetic signal strengths, strong inferences are not possible (and distributional attributes were not the intended result of their simulations). Thus, for both Pagel's $\lambda$ or Blomberg's $\kappa$, evaluation of statistical moments across a range of $\lambda$ used to generate data would be valuable for adjudicating the --relibalilty-- RELIABILITY of these statistics.  Furthermore, --these are statistics that appear to have expected values that vary with tree size-- the expected values of these statistics appear to vary with tree size [@Munkemuller_et_al2012], making comparisons across studies challenging. Therefore, transformation of these statistics into $Z$-scores would allow evaluation of the efficacy of each statistic to yield effect sizes that could be used for comparisons of the strength of phylogenetic signal across traits and lineages.

@@

**Lambda ($\lambda$) estimates of phylogenetic signal are innacurate.** Computer simulations reveal that for $\hat{\lambda}$, the distributional expectations of a Bernoulli variable were mostly upheld. First, the mean value of $\hat{\lambda}$ increases as $\lambda$ increases, but it is negatively-biased (particularly for small tree sizes), and is consistently less than the input $\lambda$ value across most of its range (Fig. 1 black line). Second, the standard deviation of $\hat{\lambda}$ is largest at intermediate values of $\lambda$ and smallest at extreme values <!-- I think the reference to (Fig.1 red line) should go here. As it is written, you'd expect the red line to represent an estimate of precision rather than SD -->, implying that the precision in estimating $\lambda$ varies across the range of input values (Fig. 1 red line). Additionally, standard deviations of $\hat{\lambda}$ are negatively associated with tree size, and for tress of 128 species or less, $\hat{\lambda}$ are quite variable, except for cases when $\lambda$ is near or equal to $1$. Third, the distributions of $\hat{\lambda}$ are not normal across its range, but become increasingly skewed at more extreme values of $\lambda$ (Fig. 1 blue line). <!-- Maybe add a quick explanation that lower values of blue line indicate higher skew. Without the explanation, a reader might glance at the figure and think "the blue line shows *lower* values at 0 and 1, but the text said it's *more* skewed at 0 and 1". I know it's in the caption, but that was my first thought when reading. --> For small tree sizes, it is also clear that distributions are more platykurtic at intermediate values of $\hat{\lambda}$. Taken together these results reveal that $\hat\lambda$ inconsistently estimates  phylogenetic signal, both across tree sizes and across the range of input values. Additional simulations (Supplemental Information) reveal that incorporating $\hat\lambda$ in PGLS anova and regression does not adversely affect the statistical properties of PGLS parameter estimation or model evaluation (type I error, power, bias in coefficients). Thus, it is reasonable to incorporate $\hat\lambda$ in PGLS as a parameter for tuning the degree of phylogenetic signal in the dependent variables during the analysis. However, the statistical properties shown in Fig. 1 demonstrate that $\lambda$ is unsuitable as an effect size for measuring the strength of phylogenetic signal in data, or for comparing that signal across datasets.

@@

**Kappa ($\kappa$) estimates of phylogenetic signal are more reliable.** Simulation results for $\hat\kappa$ demonstrate that $\hat\kappa$ displays better statistical properties. First, as expected, mean values of $\hat\kappa$ increase with increasing signal ($\lambda$) irrespective of tree size, though the increase does not scale linearly with input levels of phylogenetic signal (Fig. 2 black line). Additionally, the standard deviation of $\hat\kappa$ is consistent across tree sizes (Fig. 2 red line), and while it increases with $\lambda$, it is always less than the corresponding mean. This finding is perhaps unsurprising, as $\hat\kappa$ is lower-bounded by 0, and is never large for small values of $\lambda$. Importantly, $\hat\kappa$ is normally distributed across the range of input $\lambda$, and remains consistent in this pattern regardless of tree size (Fig. 2 blue line).  In fact, distributions of $\hat\kappa$ display slight skewing only at small tree sizes and when $\lambda$ is near or equal to $1$. This result differs from those of [@Munkemuller_et_al2012], where skewing appears to be from combining random values generated independently. These findings reveal that while $\kappa$ is more reliable as an estimate of phylogenetic signal, the non-linear scaling with input signal implies that it should not be considered an effect size that measures the strength of phylogenetic signal on a common scale for comparison across datasets.

@@

**Effect sizes from $\kappa$ ($Z_{\kappa}$) better characterize phylogenetic signal.**
To measure the strength of phylogenetic signal on a common scale, here we propose effect sizes (Z-scores) for both $\lambda$ and $\kappa$. Statistically, a standardized effect size may be found as:

@@

Here we evaluate the ability of $Z_{\lambda}$ and $Z_{\kappa}$ to characterize known levels of phylogenetic signal. Both $Z_{\lambda}$ and $Z_{\kappa}$ are associated with input phylogenetic signal ($\lambda$), indicating that both statistics capture the observed signal (Fig. 3). However, effect sizes from $\hat{\lambda}$ made little sense, as they are more strongly associated with tree size than they are with the actual phylogenetic signal in the data (Fig. 3). By contrast, $Z_{\kappa}$ is much more consistent accross tree sizes, and increases more linearly with incrEasing levels of phylogenetic signal. Additionally, $Z_{\kappa}$ exhibits a much stronger association with phylogenetic signal strength as compared to tree size (Fig. 3), and its standard deviation across input signal is more consistent. This implies that similar levels of precision are found with $Z_{\kappa}$ across the range of input values. Thus between the two statistics, $Z_{\kappa}$ is a more reliable measure of the strength of phylogenetic signal, and may be used to compare levels of phylogenetic signal across datasets.

@@

**Simulations.** Simulations were conducted by generating pure-birth phylogenies at each of six different tree sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylognetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). We generated 100 random trees for each intersection of tree size and $\lambda$. For each $\lambda$ within each tree size, continuous traits were then simulated on each phylogeny under a BM model of evolution. For each set of 100 trees we measured the mean values of $\hat{\lambda}$ and $\kappa$, their standard deviation, and calculated the Shapiro-Wilk $W$ statistic as a departure from normality (symmetry). For the latter, a value of $1.0$ indicates normally distributed values, while departures from $1.0$ indicate skewness. Simulations were then repeated for both --bananced-- BALANCED and pectinate trees, which yielded qualitatively similar results (see Supporting Information). Trees containing polytomies, and an evaluation of $\hat{\lambda}$ from models of linear regression and phylogenetic ANOVA, were also investigated, and results were qualitatively similar to those reported above (see Supporting Information).

**Empirical Data.** Surface area to volume ratios (SA:V) and relative body width ($\frac{BW}{SVL}$) measures were obtained --from as species means-- from individuals of 305 species, from which species means were obtained [@Baken2019; @Baken2020]. A time-dated molecular phylogeny for the group [@Bonett2017] was pruned to match the species in the phenotypic dataset. The phylogenetic signal in each trait was then characterized using $\kappa$, which was converted to its effect size ($Z_\kappa$) using `geomorph` 3.3.1 [@AdamsOtarola2013; @AdamsGeomorph], and routines by the authors (**to be incorporated in `geomorph` upon manuscript acceptance**).

>>>>

The appeal of Pagel's $\lambda$ and Blomberg's $\kappa$ as descriptive statistics is that they provide a basis for interpreting "weak" versus "strong" phylogenetic signal; i.e., small versus large values of $\hat{\lambda}$ or $\kappa$, respectively, in a comparative sense [@DeMeester2019; @Pintanel2019;@Su2019]. Nonetheless, an important question that has yet to be considered is whether such comparisons are analytically appropriate, and whether these statistics are, or can be, converted to effect sizes for comparative analyses across datasets. To be statistics representing phylogenetic signal, they should have reliable distributional properties, which could be revealed with simulation experiments. For instance, as a proportional random variable bounded by $0$ and $1$, we might expect that $\hat{\lambda}$ follows a Bernoulli distribution (**add ref**); i.e., branch lengths in a tree are scaled proportionally to the probability that data arise from a BM process. Given a known $\lambda$ value used to generate random data on a tree, we would also expect  that the mean of an empirical sampling distribution of $\hat{\lambda}$ would approximately equal $\lambda$; the dispersion of $\hat{\lambda}$ would be largest at intermediate values of $\lambda$, $\hat{\lambda}$ would be predictable over the range of $\lambda$ with respect to tree size; the distribution of $\hat{\lambda}$ would be symmetric at intermediate values of $\lambda$ and more skewed toward values of 0 or 1; and that the distribution of $\hat{\lambda}$ will be more platykurtic at intermediate values of $\lambda$, becoming more leptokurtic toward 0 and 1 (**add same ref**). Prior work [@Munkemuller_et_al2012] seems to support some of these conjectures, based superficially on statistical moments for a given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of ref. [@Munkemuller_et_al2012]). However, because the "strength of Brownian motion" was simulated as a varied weighted-average of data simulated on trees with $\lambda=0$ and $\lambda=1$ and not as prescribed values of $\lambda$ [@Munkemuller_et_al2012], interpretation of these patterns is challenging.

By contrast, for Blomberg's $\kappa$, which is positively unbounded, we might expect that for any $\lambda$ used to generate data, estimates of $\kappa$ might follow a normal distribution, with values distributed symmetrically about the input value. This attribute seemed less reasonable based on the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], which suggested that distributions were positively skewed and that Blomberg's $\kappa$ might not behave as a statistic that follows a normal distribution. However, because their simulations used a weighted combination of simulated phylogenetic signal strengths, strong inferences are not possible (and distributional attributes were not the intended result of their simulations). Thus, for both Pagel's $\lambda$ or Blomberg's $\kappa$, evaluation of statistical moments across a range of $\lambda$ used to generate data would be valuable for adjudicating the reliability of these statistics as effect sizes.  Furthermore, the expected values of these statistics appear to vary with tree size [@Munkemuller_et_al2012], making comparisons across studies challenging. Therefore, transformation of these statistics into $Z$-scores would allow evaluation of the efficacy of each statistic to yield effect sizes that could be used for comparisons of the strength of phylogenetic signal across traits and lineages.

@@

**Lambda ($\lambda$) estimates of phylogenetic signal are innacurate.** Computer simulations reveal that for $\hat{\lambda}$, the distributional expectations of a Bernoulli variable were mostly upheld. First, the mean value of $\hat{\lambda}$ increases as $\lambda$ increases, but it is negatively-biased (particularly for small tree sizes), and is consistently less than the input $\lambda$ value across most of its range (Fig. 1 black line). Second, the standard deviation of $\hat{\lambda}$ is largest at intermediate values of $\lambda$ and smallest at extreme values (Fig.1 red line), implying that the precision in estimating $\lambda$ varies across the range of input values. Additionally, standard deviations of $\hat{\lambda}$ are negatively associated with tree size, and for tress of 128 species or less, $\hat{\lambda}$ are quite variable, except for cases when $\lambda$ is near or equal to $1$. Third, the distributions of $\hat{\lambda}$ are not normal across its range, but become increasingly skewed at more extreme values of $\lambda$ (Fig. 1 blue line). For small tree sizes, it is also clear that distributions are more platykurtic at intermediate values of $\hat{\lambda}$. Taken together these results reveal that $\hat\lambda$ inconsistently estimates  phylogenetic signal, both across tree sizes and across the range of input values. Additional simulations (Supplemental Information) reveal that incorporating $\hat\lambda$ in PGLS anova and regression does not adversely affect the statistical properties of PGLS parameter estimation or model evaluation (type I error, power, bias in coefficients). Thus, it is reasonable to incorporate $\hat\lambda$ in PGLS as a parameter for tuning the degree of phylogenetic signal in the dependent variables during the analysis. However, the statistical properties shown in Fig. 1 demonstrate that $\lambda$ is unsuitable as an effect size for measuring the strength of phylogenetic signal in data, and thus $\lambda$ should not be used for comparing phylogenetic signal across datasets.

@@

**Kappa ($\kappa$) estimates of phylogenetic signal are more reliable.** Simulation results for $\hat\kappa$ demonstrate that $\hat\kappa$ displays better statistical properties. First, as expected, mean values of $\hat\kappa$ increase with increasing signal ($\lambda$) irrespective of tree size, though the increase does not scale linearly with input levels of phylogenetic signal (Fig. 2 black line). Additionally, the standard deviation of $\hat\kappa$ is consistent across tree sizes (Fig. 2 red line), and while it increases with $\lambda$, it is always less than the corresponding mean. This finding is perhaps unsurprising, as $\hat\kappa$ is lower-bounded by 0, and is never large for small values of $\lambda$. Importantly, $\hat\kappa$ is normally distributed across the range of input $\lambda$, and remains consistent in this pattern regardless of tree size (Fig. 2 blue line). This result differs from those of [@Munkemuller_et_al2012], where the skewing appears to be the result of combining random values generated independently, rather than being a property of $\kappa$ itself. Overall, these findings reveal that while $\kappa$ is more reliable as an estimate of phylogenetic signal, the non-linear scaling with input signal implies that it should not be considered an effect size that measures the strength of phylogenetic signal on a common scale for comparison across datasets.

@@

**Effect sizes from $\kappa$ ($Z_{\kappa}$) better characterize phylogenetic signal.**
To measure the strength of phylogenetic signal on a common scale, we propose effect sizes (Z-scores) for both $\lambda$ and $\kappa$. Statistically, a standardized effect size may be found as:

@@

Here we evaluate the ability of $Z_{\lambda}$ and $Z_{\kappa}$ to characterize known levels of phylogenetic signal. Both $Z_{\lambda}$ and $Z_{\kappa}$ are associated with input phylogenetic signal ($\lambda$), indicating that both statistics capture the observed signal (Fig. 3). However, effect sizes from $\hat{\lambda}$ made little sense, as they are more strongly associated with tree size than they are with the actual phylogenetic signal in the data (Fig. 3). By contrast, $Z_{\kappa}$ is much more consistent across tree sizes, and increases more linearly with increasing levels of phylogenetic signal. Additionally, $Z_{\kappa}$ exhibits a much stronger association with phylogenetic signal strength as compared to tree size (Fig. 3), and its standard deviation across input signal is more consistent. This implies that similar levels of precision are found with $Z_{\kappa}$ across the range of input values. Thus between the two statistics, $Z_{\kappa}$ is a more reliable measure of the strength of phylogenetic signal, and may be used to compare levels of phylogenetic signal across datasets.

@@

**Simulations.** Simulations were conducted by generating pure-birth phylogenies at each of six different tree sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylognetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). We generated 100 random trees for each intersection of tree size and $\lambda$. For each $\lambda$ within each tree size, continuous traits were then simulated on each phylogeny under a BM model of evolution. For each set of 100 trees we measured the mean values of $\hat{\lambda}$ and $\kappa$, their standard deviation, and calculated the Shapiro-Wilk $W$ statistic as a departure from normality (symmetry). For the latter, a value of $1.0$ indicates normally distributed values, while departures from $1.0$ indicate skewness. Simulations were then repeated for both balanced and pectinate trees, which yielded qualitatively similar results (see Supporting Information). Trees containing polytomies, and an evaluation of $\hat{\lambda}$ from models of linear regression and phylogenetic ANOVA, were also investigated, and results were qualitatively similar to those reported above (see Supporting Information).

**Empirical Data.** Surface area to volume ratios (SA:V) and relative body width ($\frac{BW}{SVL}$) measures were obtained from individuals of 305 species, from which species means were obtained [@Baken2019; @Baken2020]. A time-dated molecular phylogeny for the group [@Bonett2017] was pruned to match the species in the phenotypic dataset. The phylogenetic signal in each trait was then characterized using $\kappa$, which was converted to its effect size ($Z_\kappa$) using `geomorph` 3.3.1 [@AdamsOtarola2013; @AdamsGeomorph], and routines by the authors (**to be incorporated in `geomorph` upon manuscript acceptance**).
